An Axiomatic Basis for Computer Programming: Review

### 2017-02-16 hoare/paper.tex
Shortcomings and conclusion

  This is review on C.A.R. Hoare's paper, An Axiomatic Basis for Computer
  Programming.
  (WIP)

@@

``People are not perfect'', this is what we learned so fourth. 
Every one could make a mistake. In some areas it could be unnoticed, but in
others it could cause serious consequences.

@@

But more and more critical tasks entrust to computer systems. When errors and
failures cost increases, we will pay for diligence and accuracy as much as
needed.

@@

@@

In 1969 paper Hoare didn't mentioned that $Q$ should terminates, so he defined
rules just for deducting partial correctness (particular in rule D3). But later
he add definition for total correctness too.

@@

We restrict input domain as set of real numbers and according to discriminant in
predicate $P$ (suppose we used general quadratic formula):
$$ P \equiv (a, b, c \in \mathbb{R}) \wedge (b^2 - 4ac \geq 0)$$

Then we could find values by some program $Q$. Suppose as a result program
returns set of values $X$ (Don't think about implementation yet).
And finally check whether equation holds or not in R: 
$$ R \equiv (X \subseteq \mathbb{R}) \wedge (\forall x \in X \bullet ax^2 + bx + c
= 0)$$

@@

All this rules implicitly assumed the absence of side effects for $Q$ (nothing
changes except state which we define in pre and postconditions). Because in
other case, we couldn't consider correctness of overall state. \\

@@

Suppose we choose general quadratic formula for solving equation:
\begin{equation}
 x_{1,2} = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
\end{equation}

@@

Suppose we didn't see that general formula is derived from equation. Lets just
apply substitution all occurrence of $x$ in postcondition by stated expression (1).

@@

By the manipulations above we actually proved that our statement correct, with
respect to precondition $P_1$ and postcondition $R$.

And we could notice that predicate $P_1$ is stronger that predicate $P$. By
``stronger'' we mean that $P_1 \implies P$, or informally, if predicate $P_1$
holds for some values, predicate $P$ will holds either. But opposite is not true
(in our case $x = 0$ doesn't hold in $P_1$).

@@

So $Q$ will produce values which don't satisfy postcondition - program will be
incorrect according to stronger predicate.

But we could make postcondition weaker, program will still be correct.

This is actually Rule of Consequence stated in Hoare paper.\\

@@

In case when $x = 0$ (or $P \wedge \neg A$) equation takes the form:
\begin{equation}
 bx + c = 0 \text{, and } x = \frac{-c}{b}
\end{equation}

Yet again, this remains true when $B \equiv b \neq 0$ is true, so this statement
is correct according to precondition $P \wedge \neg A \wedge B$ and
postcondition $R$.

As you can see, we try discharge proof by cases, and then, using Condition
Rule (or Rule of Iteration), prove correctness of their combination.

@@

And again, two cases. Either $c = 0$ and equation holds for all
$x \in \mathbb{R}$:
\begin{equation}
  X = \mathbb{R}
\end{equation}

@@

So lets combine this four statements and prove correctness of our program.

@@

$Q_1$ correspond to equation (1), $Q_2$ to (2) and so on. $Q$ represent
nested statements:

@@

\rightline{{\rm $\Box$}}

Actually we could replace whole program with just one assignment:
$$ X = \emptyset $$

And prove it correctness according to $P$ and $R$ too, because universal
quantification expressed that every member of domain is satisfied some
predicate. This is true, when there is no members.

That's why I stressed that it doesn't help obtain right predicates.

% But sometimes it could be not so straightforward. Lets return to our example
% about real roots. Suppose quadratic coefficient a could be equals 0.

@@

% However precondition and postconditions could be almost same. This looks ugly.

% firstly we should define the
% truth.

@@

@@

@@

>>>>

  This paper is review on C.A.R. Hoare's paper, An Axiomatic Basis for Computer
  Programming \cite{hoare:69}.

  Basically an attempt is made to apply technique for reasoning about fictional
  example.

@@

``People are not perfect'', this is what we learned so fourth. 
Everyone could make a mistake. In some areas it could be unnoticed, but in
others it could cause serious consequences.

@@

But more and more critical tasks entrust to computer systems. When errors and
failures cost increases, we will pay for diligence and accuracy as much as
needed.\\

@@

Also he state axioms and rules to make inference about statements.

@@

In 1969 paper Hoare didn't mentioned that $Q$ should terminates, so he defined
rules just for deducting partial correctness (particular in rule D3). But later
he add definition for total correctness too. \\

@@

We restrict input domain as set of real numbers and according to discriminant in
predicate $P$ (suppose we used general quadratic formula and restrict domain by
cases in which equation has real roots):
$$ P \equiv (a, b, c \in \mathbb{R}) \wedge (b^2 - 4ac \geq 0)$$

Then we could find values by some program $Q$. Suppose as a result program
returns set of values $X$.
And finally check whether equation holds or not in R: 
$$ R \equiv (X \subseteq \mathbb{R}) \wedge (\forall x \in X \bullet ax^2 + bx + c
= 0)$$

@@

All this rules implicitly assumed the absence of side effects for $Q$ (nothing
changes except state which we define in pre and postconditions). Because in
other case, we couldn't consider correctness of overall state.

@@

Suppose we choose general quadratic formula for solving equation:
\begin{equation}
 X = \Big\{ \frac{-b \pm \sqrt{b^2 - 4ac}}{2a} \Big\}
\end{equation}

@@

Suppose we didn't see that general formula is directly derived from equation.
Lets just apply substitution all occurrence of $x$ in postcondition by stated
expression (1). \\

@@

By the manipulations above we actually proved that our statement correct, with
respect to precondition $P_1$ and postcondition $R$. \\

We could notice that predicate $P_1$ is stronger that predicate $P$. By
``stronger'' we mean that $P_1 \implies P$, or informally, if predicate $P_1$
holds for some values, predicate $P$ will holds either. But opposite is not true
(in our case $x = 0$ doesn't hold in $P_1$).

@@

So $Q$ will produce values which don't satisfy postcondition - program will be
incorrect according to stronger predicate.
But we could make postcondition weaker, program will still be correct.

This is Rule of Consequence, stated in Hoare paper.\\

@@

In case when $x = 0$ (or $P \wedge \neg A$) equation takes the form:
\begin{equation}
  bx + c = 0 \text{, and set } X = \Big\{ \frac{-c}{b} \Big\}
\end{equation}

Yet again, this remains true when $B \equiv b \neq 0$ is true, so this statement
is correct according to precondition $P \wedge \neg A \wedge B$ and
postcondition $R$. \\

As you can see, we try to discharge proof by cases, and then, using Condition
Rule (or Rule of Iteration), prove correctness of their combination.

@@

And again, two cases. Either $c = 0$ and equation holds for all
$x \in \mathbb{R}$:

\begin{equation}
  X = \mathbb{R}
\end{equation}

@@

So lets combine this four statements and prove correctness of our program. \\

@@

$Q_1$ correspond to equation (1), $Q_2$ to (2) and so on. $Q$ represent
nested statements, including control structures and statements $Q_2, Q_3, Q_4$.

In the same way we prove that $Q$ is correct.\\

@@

\rightline{{\rm $\Box$}}
\vspace{1em}

That's it. With one significant remark.

Actually we could replace whole program
by just one assignment: 
$$ X = \emptyset $$

And prove it correctness according to $P$ and $R$ too, because universal
quantification expressed that every member of domain is satisfied some
predicate. This is true, when there is no members. \\

\textit{That's why I stressed, that it doesn't help obtain right predicates. }
%------------------------------------------------

@@

@@

As we can see with equation example, deriving good pre and postconditions is not
trivial task. It could be ambiguous when done wrong.

The first thing that comes to mind change postcondition:
$$ R^\prime \equiv R \wedge \neg(\exists x \not\in X \bullet ax^2 + bx + c = 0)$$

To check that program returns all solutions to equation. But, in practice it
very difficult prove existential quantification's.

Also quantifiers cannot be expressed as boolean expression in ordinary
languages, which doesn't support it \cite{meyer:90}.

So defining predicates is weak part of method overall.

% But as soon as we stated right one, we could gain confidence in reliability and
% correctness of our software, as much as we believe in deductive logic.

@@

Verification software in general is difficult task.

With constantly increases complexity of software and it dependencies is hard to
tell certainly, whether particular program will accomplish task right or not.

This proof technique not only had a great influence on evolution of software
verification, but also alleviate some of the other problems which exist in
computing word.

Strict definition of subroutines' variables before and after execution, helps
understand purpose and accomplishment for potential user.

Lets not forget that Halting problem is undecidable over Turing machines.

With technique which Hoare suggests we can prove total correctness of program.

But after all, we can gain confidence in reliability and correctness of our
software as much as we believe in deductive logic, validity of our
assumptions and reliability of the electronics.

@@

\bibitem[Mey90]{meyer:90}
Bertrand Meyer.
\newblock Introduction to the Theory of Programming Languages.
\newblock {\em Prentice-Hall, Inc.}, 0-13-498510-9, 1990.
